---
title: "Reproducible Research Course Project 1"
author: "Cecilia Cruz-Ram, MD DPCOM"
date: "4/23/2018"
output: html_document
fontsize: 10pt
geometry: margin = 0.5in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

A. Get the data
```{r}
path <- setwd("/Users/sexybaboy/Documents/Files/Zetch/Online Courses/Data Science Specialization Feb18/R/Reproducible Research/Course Project 1")
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(url, file.path(path, "activity.csv"))
unzip(zipfile = "activity.csv")
```

B. Turn global options warning off
```{r}
knitr::opts_chunk$set(warning = FALSE)
```

C. Load and preprocess data
```{r}
require(ggplot2)

activity <- read.csv("./activity.csv")

activity$date <- as.Date(activity$date, "%Y-%m-%d")
weekday <- weekdays(activity$date)
activity <- as.data.frame(activity)

summary(activity)
```

```{r}
steps <- with(activity, tapply(steps, date, sum, na.rm = TRUE))

hist(steps, main = "Total # of Steps (taken per day)", xlab = "Total Steps per Day", col = "darkgreen", ylim = c(0, 20), breaks = seq(0, 25000, by = 1000))
```

D. Answers to questions

*1. What is mean total number of steps taken per day?*
  
  Mean total number of steps per day
```{r}
mean(steps)
```

  Median total number of steps per day
```{r}
median(steps)
```

*2. What is the average daily activity pattern?*
  
  Create time series plot
```{r}
ada <- aggregate(activity$steps, by = list(activity$interval), FUN = mean, na.rm = TRUE)
names(ada) <- c("interval", "mean")
plot(ada$interval, ada$mean, type = "l", col ="darkgreen", lwd = 2, xlab = "Interval", ylab = "Average # of Steps", main = "Average # of Steps (per Interval)")
```

*Which 5-minute interval, on average across all the days in the dataset, contains the maximum number   of steps?*
```{r}
ada[which.max(ada$mean), ]$interval
```

*3. Imputing missing values*
  
  Calculate and report the total number of missing values in the dataset (i.e. the total number of     rows with NAs)
  
  The following sums the NA values and reports the ratio of NA/total observations.
```{r}
require(scales)
sum(is.na(activity))
percent(sum(is.na(activity))/nrow(activity))
```

  Devise a strategy for filling in all of the missing values in the dataset. The strategy does not     need to be sophisticated. For example, you could use the mean/median for that day, or the mean for   that 5-minute interval, etc.
```{r}
imputed_steps <- ada$mean[match(activity$interval, ada$interval)]
head(activity)
tail(activity)
```

Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r}
activity_imputed <- transform(activity, steps = ifelse(is.na(activity$steps), yes = imputed_steps, no = activity$steps))
total_steps_imputed <- aggregate(steps ~ date, activity_imputed, sum)
names(total_steps_imputed) <- c("date", "daily_steps")
```

Make a histogram of the total number of steps taken each day and calculate and report the mean and median total number of steps taken per day. 
```{r}
hist(total_steps_imputed$daily_steps, col = "darkgreen", xlab = "Total Steps per Day", ylim = c(0,30), main = "Total # of Steps (taken per day)", breaks = seq(0,25000,by = 1000))
```

  Mean of the total number of steps taken per day:
```{r}
mean(total_steps_imputed$daily_steps)
```

  Median of the total number of steps taken per day:
```{r}
median(total_steps_imputed$daily_steps)
```

*Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?*

Yes. Since the NAs were replaced based on the interval means, the mean and the median for the day are now the same.

*4. Are there differences in activity patterns between weekdays and weekends?*

  Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating     whether a given date is a weekday or weekend day.
```{r}
activity$date <- as.Date(strptime(activity$date, format = "%Y-%m-%d"))
activity$datetype <- sapply(activity$date, function(x) {
        if (weekdays(x) == "Saturday" | weekdays(x) == "Sunday") 
                {y <- "Weekend"} else 
                {y <- "Weekday"}
                y
        })
```

  Make a panel plot containing a time series plot (i.e. type = “l”) of the 5-minute interval (x-axis)   and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).    See the README file in the GitHub repository to see an example of what this plot should look like    using simulated data.
```{r}
activity_by_date <- aggregate(steps~interval + datetype, activity, mean, na.rm = TRUE)
plot<- ggplot(activity_by_date, aes(x = interval , y = steps, color = datetype)) +
       geom_line() +
       labs(title = "Average Daily Steps (by type of date)", x = "Interval", y = "Average # of Steps") +
       facet_wrap(~datetype, ncol = 1, nrow = 2)
print(plot)
```